Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2022-10-15 analysis/paper/paper.qmd
finish editing background sections

## Why Stemmed Points?

Stemmed points (Sumbejjirugae) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two arises that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made). Slight retouch is located on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft. These types of artifacts are often called tanged points, but we prefer 'stemmed point' here to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea [@park2022did]. After the first discovery at the Suyanggae site, more than 450 stemmed points have been found in over 30 sites [@Chong_2021]. The emergence of stemmed points is relevant to questions about the origins of projectile technology, the emergence of the Late Paleoltihic, and relationships to adjacent regions in East Asia, such as Japan.

The stemmed point is the first composite tool in the Korean Paleolithic that requires two different parts, consisting of a stone point and shaft, presumably made out of wood [@seong2008tanged]. Using blades enables mass production for this composite tool and its shape becomes more standardized [@lee2015characteristics; @park2022did]. Therefore, @o2018origins claims that understanding the emergence of projectile technology gives insights to greater cultural, evolutionary, and behavioral cognitive flexibility.

Since the stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origin(s) becomes a key for understanding the technological transition from the Early Paleoltihic, modern human dispersals, and existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern]. There has been a northworthy debate about explaining the origin(s) of stemmed points and it can be summarized into two competing models: 'heterogenic' migration [@bae2010origin], in situ evolution [@seong2009emergence]. The migration model claims that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins (e.g. 1: North route: Siberia, Mongolia, or other regions of northeast China, and 2: South route: southern China) as the result of continuing influx of modern human migration from two routes. The 'in situ' model argues that stemmed points and other Late Paleolithic technologies, including blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence. The difference between the two models comes from how to understand a few early sites that contain stemmed points but no blades and blade coares. The 'in situ' model claims that the stemmed points appeared before blades by acknowledging those early sites including Bonggok, Songamri, Yonghodong, and Hwadaeri sites while migration model supports the traditional definition of the Late or Upper Palaeolithic in Europe and Africa that started with emergence and establishment of blade technology [@seong2016eastern].

Stemmed points are important proxies to understand human behaviors during the Late Paleolithic. @park2022did examine mobility and site occupation patterns through lithic assemblages and concepts of human behavioral ecology and find that forager groups using stemmed points may have been associated with occupation of marginal or extreme environments compared to the groups with no stemmed points. Also, they were more associated with expedient technologies, indicating residential and less mobile behaviors. @Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent the characteristics of occupations such as "limited activity station" and "residential base camp." For example, stemmed points with high morphological variations in a tool size, shape of edge, degrees of damage, and types of edge retouching from Yongsandong site indicate that the site was used for specific or limited activities such as hunting [@kim2004yongsan; @seong2015diversity; @Chong_2021].

The presence of stemmed points in both Korean and Japan has been studied since the late 1980s as a part of evidence for long-distance/maritime cultural interchanges or social networks [@lee2015characteristics]. Stemmed points from the Bonggok site in Korea are currently accepted as the oldest ones among Northeast Asia, dated to ca. 41.5 ka, and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin]. After their first appearance in Korea, stemmed points (Hakuhensenntouki in Japanese) appear in Kyushu, Japan during late MIS 3. In addition to the stemmed points, there are similar artifacts found in both regions such as microblade cores, Moppule-seokgi (Kakusuijyosekki in Japanese), backed knife, bilateral points, bifacial points, and transport of obsidian [@lee2012characteristics; @chang2013human; @lee2015characteristics; @kim2021evidence]. In addition, stemmed points also help understand the connections and technological transitions between islands in Japan [@chang2013human].

@@

## Previous studies about the stemmed points function

Though the stemmed points are generally assumed as hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to understand the function of the tools without knowing their complete shape attached to other components. Preserved wooden components of projectile tools are too few and rare to generalize the overall scale and variability of the tools used in prehistory [@shea2006origins]. @Lee_Kong_2002 claimed to name the stemmed points as 'stemmed tools' because of the diverse shape of stemmed points, which could represent different functions as well as the uncertainty of complete tool shape. Other researchers agree that some stemmed points that are not symmetrical and have retouch on one side of the artifacts or have denticulate blades on one side should not be called as stemmed points but stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] @fig-sp-hand-made: 7-8.

Most stemmed points are symmetrical from the tip to the tang, having the central axis function as a line of symmetry [@Lee_Jang_2011]. The studies of the stemmed points manufacturing process and the patterns of broken pieces show that stemmed points might be used mainly as spear tips [@Lee_1985; @Chang_2002]. There are a high proportion of broken tips and stems and the reused tools were repaired based on keeping the symmetry [@Park_2013; @Kim]. In addition to the morphological aspect of stemmed points, investigating the whole site and tool composition of an assemblage show that stemmed points or stemmed tools are strongly associated with hunting activities including peeling the animal skin after slaughtering or separating the bones from the flesh [@seong2008tanged; @Chong_2021].

A recent study conducts use-wear analysis on 95 stemmed points from the Jingeuneul site along with fracture patterns, TCSA, tip cross-sectional perimeter (TCSP), and neck width to understand the use of the tools [@lee2019were]. Their results show that a considerable number of the stemmed points show diagnostic impact fractures (DIFs) on the surface, which could occur due to longitudinal force from the shaft when a stone tip hits a target. In other words, their research indicates that stemmed points could be used as spear-throwers or bows [@sano2015backed].

@@

The tip cross-sectional area (TCSA) has been used as a ballistically significant standard to discriminate different functional classes of projectile armatures such as spearthrower (a.k.a. atlatl) dart tips, arrowheads and large experimental stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins]. TCSA metric indicates the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by @hughes1998getting and and tested by @shea2006origins through comparing archeological examples to ethnographically collected samples of known use. @shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe and claimed that projectile weapons first appeared in Africa. The main advantage of the TCSA is, regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 × maximum width × maximum thickness) [@sisk2011african; @lombard2020tip]. Later, @sisk2011african proposed an alternate measure, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect. However, TCSP has a few disadvantages to directly apply to our case of stemmed points. The force and penetration depth are not only affected by the stone tip but also by the mass of the shaft, which cannot be known for archaeological stone-tipped weapons [@lombard2020tip]. @sisk2011 also mentioned that TCSP cannot be applied to backed pieces that are hafted as projectile armatures.

@lee2019were first applied TCSA to stemmed points from Korea along with their use-wear analysis. The stemmed points for their research were discovered from Jingeuneul site located in southwest of Korea, where the largest number of stemmed points (n = 99) were found in a single site. For TCSA, they were only able to use ten stemmed points because they needed to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and displayed diagnostic impact fractures for the use-wear analysis. The purpose of using TCSA in their research is to compare the values to North American dart tips and arrowheads. Their results show that the TCSA range of Jingeuneul stemmed points is relatively wide overlapped with both North American dart tips and arrowheads. Inspired by @lee2019were in addition to other prior research that developed and applied TCSA into archaeological materials from other parts of the world, we explore stemmed points from all over South Korea to better understand their function to understand the role that stemmed points represent during the Late Paleolithic.

>>>>

## Stemmed Points in Korea and East Asia

Stemmed points (Sumbejjirugae) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two arises that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made). Slight retouch is typically located on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft. Elsewhere in the world these types of artifacts are often called tanged points, but we prefer 'stemmed point' here to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea [@park2022did]. After the first discovery at the Suyanggae site, more than 450 stemmed points have been found in over 30 sites throughout South Korea [@Chong_2021]. Understanding the emergence of stemmed points is relevant to general questions about the origins of projectile technology, the emergence of the Late Palaeolithic, and relationships to adjacent regions in East Asia, such as Japan.

The stemmed point is the first composite tool in the Korean Paleolithic that requires two different parts, consisting of a stone point and shaft, presumably made out of wood [@seong2008tanged]. Using blades as the blank for the point enables mass production for this composite tool and its shape can become more standardized [@lee2015characteristics; @park2022did]. Therefore, @o2018origins claims that understanding the emergence of projectile technology gives insights to greater cultural, evolutionary, and behavioral cognitive flexibility.

Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origin(s) becomes a key for understanding the technological transition from the Early Palaeolithic, modern human dispersals, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern]. There has been a noteworthy debate about explaining the origin(s) of stemmed points which can be summarized into two competing models: a 'heterogenic' migration [@bae2010origin], and an *in situ* evolution model [@seong2009emergence]. The migration model claims that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins (e.g. 1: North route: Siberia, Mongolia, or other regions of northeast China, and 2: South route: southern China) as the result of a continuing influx of modern human migration from two routes. The *in situ* model argues that stemmed points and other Late Paleolithic technologies, including blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence. The difference between the two models comes from how to understand a few early sites that contain stemmed points but no blades and blade cores. The *in situ* model claims that stemmed points appeared before blades by acknowledging those early sites, including Bonggok, Songamri, Yonghodong, and Hwadaeri, while the migration model supports the traditional definition of the Late or Upper Palaeolithic in Europe and Africa that started with emergence and establishment of blade technology [@seong2016eastern].

Stemmed points are important proxies to understand human behaviors during the Late Palaeolithic. @park2022did examined mobility and site occupation patterns by applying concepts of human behavioral ecology to lithic assemblages and found that forager groups using stemmed points may have been associated with the occupation of marginal or extreme environments, compared to groups with no stemmed points. Also, stemmed points were more frequently associated with expedient technologies, indicating residential and less mobile behaviors. @Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent specific characteristics of occupation such as a "limited activity station" and a "residential base camp." For example, stemmed points with high morphological variations in a tool size, shape of edge, degrees of damage, and types of edge retouching from Yongsandong site may indicate that the site was used for specific or limited activities such as hunting [@kim2004yongsan; @seong2015diversity; @Chong_2021].

Stemmed points in both Korea and Japan have been studied since the late 1980s as a part of evidence for long-distance/maritime cultural interchanges or social networks [@lee2015characteristics]. Stemmed points from the Bonggok site in Korea are currently accepted as the oldest ones among Northeast Asia, dated to ca. 41.5 ka, and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin]. After their first appearance in Korea, stemmed points (Hakuhensenntouki in Japanese) appear in Kyushu, Japan during late MIS 3. In addition to the stemmed points, there are similar artifacts found in both regions such as microblade cores, Moppule-seokgi (Kakusuijyosekki in Japanese), backed knife, bilateral points, bifacial points, and transport of obsidian [@lee2012characteristics; @chang2013human; @lee2015characteristics; @kim2021evidence]. Stemmed points are thus important for understanding the connections and technological transitions between islands in Japan [@chang2013human].

@@

## Uses of stemmed points

Though stemmed points are generally assumed to have been hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to understand the likely uses of the tools without knowing their complete shape when attached to other components. Preserved wooden components of projectile tools are too few and rare to generalize the overall scale and variability of the tools used in prehistory [@shea2006origins]. @Lee_Kong_2002 claimed stemmed points were 'stemmed tools' because of their diverse shapes, which could represent different functions, as well as the uncertainty of complete tool shape. Other researchers agree that some stemmed points that are not symmetrical and have retouch on one side of the artifacts or have denticulate blades on one side should not be called stemmed points but instead stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] @fig-sp-hand-made: 7-8.

Most stemmed points are symmetrical from the tip to the tang, having the central axis function as a line of symmetry [@Lee_Jang_2011]. Studies of stemmed point manufacturing processes and the patterns of broken pieces show that stemmed points may have been used mainly as spear tips [@Lee_1985; @Chang_2002]. There are a high proportion of broken tips and stems and the reused tools were repaired based on keeping the symmetry [@Park_2013; @Kim]. In addition to the morphological aspect of stemmed points, investigation of the whole site and tool composition of an assemblage shows that stemmed points or stemmed tools are strongly associated with hunting activities including peeling the animal skin after slaughtering or separating the bones from the flesh [@seong2008tanged; @Chong_2021].

A recent study conducted use-wear analysis on 95 stemmed points from the Jingeuneul site, along with examining fracture patterns, TCSA, tip cross-sectional perimeter (TCSP), and neck width to understand the use of the tools [@lee2019were]. Their results show that a considerable number of the stemmed points show diagnostic impact fractures (DIFs) on the surface, which likely occurred due to longitudinal forces from the shaft when a stone tip hits a target. In other words, their research indicates that stemmed points may have been used as spear-throwers or bows [@sano2015backed].

@@

Tip cross-sectional area (TCSA) of stone artefacts has been used as a ballistically significant standard to discriminate different functional classes of projectile armatures such as spearthrower (a.k.a. atlatl) dart tips, arrowheads and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins]. The TCSA metric indicates the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by @hughes1998getting and and tested by @shea2006origins by comparing archaeological examples to ethnographically collected samples of known use. @shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, and claimed that projectile weapons first appeared in Africa. A key advantage of the TCSA metric is its convenience: regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 × maximum width × maximum thickness) [@sisk2011african; @lombard2020tip]. Later, @sisk2011african proposed an alternate measure, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect. However, TCSP has a few disadvantages that limit its applicability to our case study of Korean stemmed points. The force and penetration depth are not only affected by the stone tip, but also by the mass of the shaft, which cannot be known for most archaeological stone-tipped weapons [@lombard2020tip]. @sisk2011 also mentioned that TCSP cannot be applied to backed pieces that are hafted as projectile armatures.

@lee2019were first applied TCSA to stemmed points from Korea along with use-wear analysis. In that study, stemmed points were analysed from Jingeuneul, located in southwest of Korea, which has the largest number of stemmed points (n = 99) ever found in a single site. For TCSA, they were only able to use ten stemmed points because they needed to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and also displayed diagnostic impact fractures for the use-wear analysis. The purpose of using TCSA in their research was to compare the values to North American dart tips and arrowheads. Their results show that the TCSA range of Jingeuneul stemmed points is relatively wide, overlapping with both North American dart tips and arrowheads. Inspired by @lee2019were and TCSA research from lithic assemblages in other parts of the world, we aimed to investigate TCSA values for a large number of stemmed points from all over South Korea to better understand their likely uses during the Late Paleolithic.
